# 2024-03-08 Experiment analysis

## Dependencies and setup

```{r}
library(tidyverse)
library(ggplot2)
library(cowplot)
library(RColorBrewer)
library(khroma)
library(rstatix)
library(knitr)
library(kableExtra)
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R")
```

```{r}
analysis_path <- "experiments/2024-03-08-varied-interaction-matrices/analysis"
data_path <- paste(analysis_path, "data", "world_summary.csv", sep = "/")
data <- read_csv(data_path)
```

## Data preprocessing

```{r}
data <- data %>%
  mutate(
    interaction_matrix = as.factor(interaction_matrix),
    graph_type = as.factor(graph_type),
    summary_mode = as.factor(summary_mode),
    update = as.numeric(update),
    SEED = as.factor(SEED)
  )

# Separate proof-of-concept runs from other interaction matrics
poc_data <- data %>% filter(interaction_matrix == "orig-pof")
data <- data %>%
  filter(interaction_matrix != "orig-pof") %>%
  mutate(
    im_connectance = case_when(
      str_detect(interaction_matrix, "c25") ~ "25",
      str_detect(interaction_matrix, "c50") ~ "50",
      str_detect(interaction_matrix, "c75") ~ "75"
    ),
    im_pip = case_when(
      str_detect(interaction_matrix, "pip25") ~ "25",
      str_detect(interaction_matrix, "pip50") ~ "50",
      str_detect(interaction_matrix, "pip75") ~ "75"
    )
  ) %>%
  mutate(
    im_connectance = as.factor(im_connectance),
    im_pip = as.factor(im_pip)
  )

max_update <- max(data$update)
final_update_data <- data %>%
  filter(update == max_update)
# write_csv(final_update_data, paste(analysis_path, "data", "world_summary_final_update.csv", sep="/"))

```

```{r}
theme_set(theme_cowplot())
```

```{r}
plot_time_series <- function(plt_data, summary_md, interaction_mat) {

  pd <- plt_data %>% filter(
    summary_mode == summary_md & interaction_matrix == interaction_mat
  )

  plot <- pd %>%
    ggplot(
      aes(
        x = update,
        y = logged_mult_score,
        color = graph_type,
        fill = graph_type
      )
    ) +
    stat_summary(geom="line", fun="mean") +
    stat_summary(
      geom="ribbon",
      fun.data="mean_cl_boot",
      fun.args=list(conf.int=0.95),
      alpha=0.2,
      linetype=0
    ) +
    labs(title=interaction_mat)

  return(plot)
}
```

## Visualize transitionability scores over time

First, `logged_mult_score`:

```{r}

data <- data %>% filter((update %% 100) == 0)

plot <- plot_time_series(data, "ranked_threshold", "c25pip25")
# plot
ggsave(
  paste(analysis_path, "plots", "c25pip25_ranked_thresh_logged_mult_score.pdf", sep="/"),
  plot = plot
)

plot <- plot_time_series(data, "ranked_threshold", "c25pip50")
# plot
ggsave(
  paste(analysis_path, "plots", "c25pip50_ranked_thresh_logged_mult_score.pdf", sep="/"),
  plot = plot
)

plot <- plot_time_series(data, "ranked_threshold", "c25pip75")
# plot
ggsave(
  paste(analysis_path, "plots", "c25pip75_ranked_thresh_logged_mult_score.pdf", sep="/"),
  plot = plot
)
```

```{r}
plot <- plot_time_series(data, "ranked_threshold", "c50pip25")
# plot
ggsave(
  paste(analysis_path, "plots", "c50pip25_ranked_thresh_logged_mult_score.pdf", sep="/"),
  plot = plot
)

plot <- plot_time_series(data, "ranked_threshold", "c50pip50")
# plot
ggsave(
  paste(analysis_path, "plots", "c50pip50_ranked_thresh_logged_mult_score.pdf", sep="/"),
  plot = plot
)

plot <- plot_time_series(data, "ranked_threshold", "c50pip75")
# plot
ggsave(
  paste(analysis_path, "plots", "c50pip75_ranked_thresh_logged_mult_score.pdf", sep="/"),
  plot = plot
)
```

```{r}
plot <- plot_time_series(data, "ranked_threshold", "c75pip25")
# plot
ggsave(
  paste(analysis_path, "plots", "c75pip25_ranked_thresh_logged_mult_score.pdf", sep="/"),
  plot = plot
)

plot <- plot_time_series(data, "ranked_threshold", "c75pip50")
# plot
ggsave(
  paste(analysis_path, "plots", "c75pip50_ranked_thresh_logged_mult_score.pdf", sep="/"),
  plot = plot
)

plot <- plot_time_series(data, "ranked_threshold", "c75pip75")
# plot
ggsave(
  paste(analysis_path, "plots", "c75pip75_ranked_thresh_logged_mult_score.pdf", sep="/"),
  plot = plot
)
```

```{r}
plot <- plot_time_series(filter(poc_data, (update%%100)==0), "ranked_threshold", "orig-pof")
# plot
ggsave(
  paste(analysis_path, "plots", "poc_ranked_thresh_logged_mult_score.pdf", sep="/"),
  plot = plot + ggtitle("POC")
)



summary(filter(poc_data, update==max_update & graph_type == "star" & summary_mode == "ranked_threshold")$logged_mult_score)
```

### Final update

```{r}
library(ggh4x)
rt_final_data <- final_update_data %>%
  filter(summary_mode == "ranked_threshold")

```

```{r}

rt_final_data <- rt_final_data %>%
  mutate(
    Connectance = case_when(
      im_connectance == "25" ~ "0.25",
      im_connectance == "50" ~ "0.50",
      im_connectance == "75" ~ "0.75"
    ),
    PIP = case_when(
      im_pip == "25" ~ "0.25",
      im_pip == "50" ~ "0.50",
      im_pip == "75" ~ "0.75"
    )
  )

wm_median <- rt_final_data %>%
  filter(graph_type == "well-mixed") %>%
  # group_by(interaction_matrix, im_connectance, im_pip) %>%
  group_by(interaction_matrix, Connectance, PIP) %>%
  dplyr::summarize(wm_median = median(logged_mult_score))

graph_ticks <- c("well-mixed", "toroidal-lattice", "linear-chain", "cycle", "wheel", "star", "windmill", "comet-kite", "random-barabasi-albert", "random-waxman")
graph_labels <- c("Well mixed", "Toroidal lattice", "Linear chain", "Cycle", "Wheel", "Star", "Windmill", "Comet-kite", "Barabasi-Albert", "Waxman")

plot_final <- ggplot(
    rt_final_data,
    aes(
      x = graph_type,
      y = logged_mult_score,
      fill = graph_type
    )
  ) +
  # geom_flat_violin(
  #   position = position_nudge(x = .2, y = 0),
  #   alpha = .8,
  #   adjust = 1.5
  # ) +
  geom_point(
    mapping = aes(color = graph_type),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    # width = .1,
    outlier.shape = NA,
    alpha = 0.5
  ) +
  geom_hline(
    data = wm_median,
    aes(yintercept = wm_median),
    linetype = "dashed"
  ) +
  scale_color_brewer(palette = "Set3") +
  scale_fill_brewer(palette = "Set3") +
  scale_x_discrete(
    name = "Spatial structure",
    limits = graph_ticks,
    breaks = graph_ticks,
    labels = graph_labels
  ) +
  scale_y_continuous(
    name = "Community Transitionability"
  ) +
  ggh4x::facet_grid2(
    Connectance ~ PIP,
    labeller = label_both
    # independent = "y"
  ) +
  coord_flip() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    panel.border = element_rect(color = "gray", size = 2)
  )
# plot_final
ggsave(
  paste(
    analysis_path,
    "plots",
    "final_ranked_thresh_logged_mult_score.pdf",
    sep = "/"
  ),
  plot = plot_final,
  width = 8.5,
  height = 8
)
```

<!-- TODO - stats -->

For reference, a table of means and medians.

```{r}
summary_table <- rt_final_data %>%
  dplyr::group_by(interaction_matrix, graph_type) %>%
  dplyr::summarize(
    score_median = median(logged_mult_score),
    score_mean = mean(logged_mult_score),
    replicates = n()
  ) %>%
  arrange(score_median, .by_group = TRUE) %>%
  kable() %>%
  kable_styling(
    latex_options = "striped"
  )
summary_table
# summary_table

save_kable(
  summary_table,
  paste(
    analysis_path,
    "plots",
    paste0("summary_table.pdf"),
    sep = "/"
  )
)
```


First, a Kruskal-Wallis test (per-interaction matrix) to test for significant differences in distributions across spatial structures.

```{r}
kw_test <- rt_final_data %>%
  group_by(interaction_matrix) %>%
  kruskal_test(logged_mult_score ~ graph_type) %>%
  mutate(sig = (p < 0.05))

kable(kw_test)
# kw_test
```

Next, perform pairwise wilcoxon rank-sum tests for all significant comparison groups.

```{r}
# Grab group names of significant comparisons
sig_kw_groups <- filter(kw_test, p < 0.05)$interaction_matrix

wrs_test <- rt_final_data %>%
  filter(
    interaction_matrix %in% sig_kw_groups
  ) %>%
  group_by(interaction_matrix) %>%
  pairwise_wilcox_test(logged_mult_score ~ graph_type) %>%
  adjust_pvalue(method = "holm") %>%
  add_significance("p.adj")

wrs_test_table <- kable(wrs_test) %>%
  kable_styling(
    latex_options = "striped"
  )
wrs_test_table

save_kable(
  wrs_test_table,
  paste(
    analysis_path,
    "plots",
    paste0("wrs_test_results.pdf"),
    sep = "/"
  )
)
```

Filter pairwise tests to just well-mixed.

```{r}
wm_wrs_test_table <- wrs_test %>%
  filter(group1 == "well-mixed" | group2 == "well-mixed") %>%
  kable() %>%
  kable_styling(
    latex_options = "striped"
  )

save_kable(
  wm_wrs_test_table,
  paste(
    analysis_path,
    "plots",
    paste0("wm_wrs_test_results.pdf"),
    sep = "/"
  )
)
```