# Community transitionability analyses

## Dependencies and setup

```{r}
library(tidyverse)
library(cowplot)
library(RColorBrewer)
library(khroma)
library(rstatix)
library(knitr)
library(kableExtra)
library(ggh4x)
source("https://gist.githubusercontent.com/benmarwick/2a1bb0133ff568cbe28d/raw/fb53bd97121f7f9ce947837ef1a4c65a73bffb3f/geom_flat_violin.R")
```

```{r}
# Check if Rmd is being compiled using bookdown
bookdown <- exists("bookdown_build")
```

```{r}
experiment_slug <- "2024-03-08-varied-interaction-matrices"
working_directory <- paste(
  "experiments",
  experiment_slug,
  "analysis",
  sep = "/"
)
# Adjust working directory if being knitted for bookdown build.
if (bookdown) {
  working_directory <- paste0(
    bookdown_wd_prefix,
    working_directory
  )
}

plot_dir <- paste(
  working_directory,
  "plots",
  sep = "/"
)

# Load summary data from final update
data_path <- paste(
  working_directory,
  "data",
  "world_summary_final_update.csv",
  sep = "/"
)
data <- read_csv(data_path)
```

Set cowplot theme as default plotting theme.

```{r}
theme_set(theme_cowplot())
```

## Data preprocessing

```{r}
data <- data %>%
  mutate(
    interaction_matrix = as.factor(interaction_matrix),
    graph_type = as.factor(graph_type),
    summary_mode = as.factor(summary_mode),
    update = as.numeric(update),
    SEED = as.factor(SEED)
  )

# Separate proof-of-concept runs from other interaction matrics
# (we don't use the proof-of-concept in our analyses)
poc_data <- data %>% filter(interaction_matrix == "orig-pof")
data <- data %>%
  filter(interaction_matrix != "orig-pof") %>%
  mutate(
    im_connectance = case_when(
      str_detect(interaction_matrix, "c25") ~ "25",
      str_detect(interaction_matrix, "c50") ~ "50",
      str_detect(interaction_matrix, "c75") ~ "75"
    ),
    im_pip = case_when(
      str_detect(interaction_matrix, "pip25") ~ "25",
      str_detect(interaction_matrix, "pip50") ~ "50",
      str_detect(interaction_matrix, "pip75") ~ "75"
    )
  ) %>%
  mutate(
    im_connectance = as.factor(im_connectance),
    im_pip = as.factor(im_pip)
  )

# Ensure that we're isolating values from end-of-simulation.
max_update <- max(data$update)
final_update_data <- data %>%
  filter(update == max_update)
```

There are several different summarization methods supported by the chemical ecology model software.
Here, we use the ranked-threshold metric.

```{r}
rt_final_data <- final_update_data %>%
  filter(summary_mode == "ranked_threshold")

rt_final_data <- rt_final_data %>%
  mutate(
    Connectance = case_when(
      im_connectance == "25" ~ "0.25",
      im_connectance == "50" ~ "0.50",
      im_connectance == "75" ~ "0.75"
    ),
    PIP = case_when(
      im_pip == "25" ~ "0.25",
      im_pip == "50" ~ "0.50",
      im_pip == "75" ~ "0.75"
    )
  )
```

Calculate the median transitionability score (`logged_mult_score`) for each well-mixed regime.

```{r}
wm_median <- rt_final_data %>%
  filter(graph_type == "well-mixed") %>%
  dplyr::group_by(interaction_matrix, Connectance, PIP) %>%
  dplyr::summarize(wm_median = median(logged_mult_score))
```

## Final transitionability scores

We visualize the final community transitionability scores (`logged_mult_score`) for each spatial structure regime across for each interaction network.
For each interaction matric, we draw a vertical dashed line (black) to indicate median transitionability score achived in the well-mixed regime.
This value serves as the baseline expectation in the absence of spatial structure.

Additionally, we draw a solid vertical line (red) to indicate 0 on the transitionability score axis.
Transitionability scores greater than zero indicate that a community exhibited dynamics more closely resembling pure adaptive dynamics than pure ecological dynamics.
Transitionability scores less than zero indicate that a community exhibited dynamics more closely resembling pure ecological dynamics than pure adaptive dynamics.

```{r}
# Provide explicit ordering for graph ticks/labels
graph_ticks <- c(
  "well-mixed",
  "toroidal-lattice",
  "linear-chain",
  "cycle",
  "wheel",
  "star",
  "windmill",
  "comet-kite",
  "random-barabasi-albert",
  "random-waxman"
)
graph_labels <- c(
  "Well mixed",
  "Toroidal lattice",
  "Linear chain",
  "Cycle",
  "Wheel",
  "Star",
  "Windmill",
  "Comet-kite",
  "Barabasi-Albert",
  "Waxman"
)

plot_final <- ggplot(
    rt_final_data,
    aes(
      x = graph_type,
      y = logged_mult_score,
      fill = graph_type
    )
  ) +
  geom_hline(
    yintercept = 0,
    color = "red",
    linetype = "solid",
    alpha = 0.65
  ) +
  geom_point(
    mapping = aes(color = graph_type),
    position = position_jitter(width = .15),
    size = .5,
    alpha = 0.8
  ) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.5
  ) +
  geom_hline(
    data = wm_median,
    aes(yintercept = wm_median),
    linetype = "dashed"
  ) +
  scale_color_brewer(palette = "Set3") +
  scale_fill_brewer(palette = "Set3") +
  scale_x_discrete(
    name = "Spatial structure",
    limits = graph_ticks,
    breaks = graph_ticks,
    labels = graph_labels
  ) +
  scale_y_continuous(
    name = "Community Transitionability"
  ) +
  ggh4x::facet_grid2(
    Connectance ~ PIP,
    labeller = label_both
  ) +
  coord_flip() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 30,
      hjust = 1
    ),
    panel.border = element_rect(color = "gray", size = 2)
  )

ggsave(
  paste(
    plot_dir,
    "final_ranked_thresh_logged_mult_score.pdf",
    sep = "/"
  ),
  plot = plot_final,
  width = 8.5,
  height = 8
)

plot_final
```

For reference, we generate a table of mean and median transitionability scores per-regime per-experiment:

```{r}
summary_data <- rt_final_data %>%
  dplyr::group_by(interaction_matrix, graph_type) %>%
  dplyr::summarize(
    score_median = median(logged_mult_score),
    score_mean = mean(logged_mult_score),
    replicates = n()
  ) %>%
  arrange(score_median, .by_group = TRUE)

summary_table <- summary_data %>%
  kable() %>%
  kable_styling(
    latex_options = "striped"
  )
save_kable(
  summary_table,
  paste(
    plot_dir,
    "summary_table.pdf",
    sep = "/"
  )
)
summary_table
```

### Statistical analyses

#### Transitionabiliy score distributions - Kruskal-Wallis test results

First, a Kruskal-Wallis test (per-interaction matrix) to test for significant differences in distributions across spatial structure regimes.

```{r}
kw_test <- rt_final_data %>%
  group_by(interaction_matrix) %>%
  kruskal_test(logged_mult_score ~ graph_type) %>%
  mutate(sig = (p < 0.05))

kw_table <- kw_test %>%
  kable() %>%
  kable_styling(
    latex_options = "striped"
  )
save_kable(
  kw_table,
  paste(
    plot_dir,
    "kw_test_results.pdf",
    sep = "/"
  )
)
kw_table
```

#### Transitionability score distributions - Pairwise Wilcoxon rank-sum test results

Next, we perform pairwise Wilcoxon rank-sum tests for all significant comparison groups.
We use a Holm-Bonferroni correction for multiple comparisons.

```{r}
# Grab group names of significant comparisons
sig_kw_groups <- filter(kw_test, p < 0.05)$interaction_matrix
# Perform pairwise rank-sum tests, adjust for multiple comparisons
wrs_test <- rt_final_data %>%
  filter(
    interaction_matrix %in% sig_kw_groups
  ) %>%
  group_by(interaction_matrix) %>%
  pairwise_wilcox_test(logged_mult_score ~ graph_type) %>%
  adjust_pvalue(method = "holm") %>%
  add_significance("p.adj")
# Build a pretty table
wrs_test_table <- kable(wrs_test) %>%
  kable_styling(
    latex_options = "striped"
  )
save_kable(
  wrs_test_table,
  paste(
    plot_dir,
    "wrs_test_results.pdf",
    sep = "/"
  )
)
wrs_test_table
```

## Identify amplifiers or suppressors

Next, we categorize each spatial structure as an amplifier or suppressor based on
its transitionability score relative to the well-mixed regime.
Spatial structure regimes with scores that are greater than well-mixed (statistically significant) are categorized as amplifiers.
Spatial structure regimes with scores that are lower than well-mixed (statistically significant) are categorized as suppressors.
If we failed to detect a statistically significant difference between a spatial structure and well-mixed, we categorized it as "neither".

First, filter the pairwise tests to just those that involve the well-mixed regime.

```{r}
wm_wrs_test_table <- wrs_test %>%
  filter(group1 == "well-mixed" | group2 == "well-mixed") %>%
  kable() %>%
  kable_styling(
    latex_options = "striped"
  )
save_kable(
  wm_wrs_test_table,
  paste(
    plot_dir,
    "wm_wrs_test_results.pdf",
    sep = "/"
  )
)
wm_wrs_test_table
```

For each spatial structure, identify amplifiers and suppressors.

```{r}
int_matrices <- unique(as.character(wrs_test$interaction_matrix))
well_mixed_comps <- wrs_test %>%
  filter(group1 == "well-mixed" | group2 == "well-mixed") %>%
  mutate(
    non_wm_graph = case_when(
      group1 == "well-mixed" ~ group2,
      group2 == "well-mixed" ~ group1
    )
  )
non_wm_graph_types <- unique(as.character(well_mixed_comps$non_wm_graph))

spatial_struct_effects <- data.frame(
  interaction_matrix = character(),
  graph_type = character(),
  effect = character(),
  wm_median_score = numeric(),
  graph_median_score = numeric(),
  sig = logical()
)

# Identify promotors (significant and > well-mixed)
# Identify represssors (significant and < well-mixed)
# Neither (not significant)
# The output of this loop is sanity-checked against statistical results table.
for (interaction_mat in int_matrices) {
  # Get median score for well-mixed
  wm_median_score <- filter(
    summary_data,
    graph_type == "well-mixed" & interaction_matrix == interaction_mat
  )$score_median[[1]]
  # Get relevent wilcoxon rank-sum comparisons
  im_comps <- well_mixed_comps %>%
    filter(interaction_matrix == interaction_mat)
  for (graph in non_wm_graph_types) {
    graph_median_score <- filter(
      summary_data,
      graph_type == graph & interaction_matrix == interaction_mat
    )$score_median[[1]]
    comp_info <- filter(im_comps, non_wm_graph == graph)
    sig <- comp_info$p.adj[[1]] < 0.05
    effect <- "unknown"
    if (sig && graph_median_score < wm_median_score) {
      effect <- "suppressor"
    } else if (sig && graph_median_score > wm_median_score) {
      effect <- "promoter"
    } else {
      effect <- "neither"
    }
    spatial_struct_effects <- add_row(
      spatial_struct_effects,
      interaction_matrix = interaction_mat,
      graph_type = graph,
      effect = effect,
      wm_median_score = wm_median_score,
      graph_median_score = graph_median_score,
      sig = sig
    )
  }
}

effect_table <- spatial_struct_effects %>%
  kable() %>%
  kable_styling(
    latex_options = "striped"
  )

save_kable(
  effect_table,
  paste(
    plot_dir,
    "spatial_struct_effect_table.pdf",
    sep = "/"
  )
)

effect_table
```

Break effects down by interaction matrix (experiment).
Arrange in order of effect-size.

```{r}
for (im in int_matrices) {
  max_promoter <- max(
    filter(
      spatial_struct_effects,
      interaction_matrix == im & effect == "promoter"
    )$graph_median_score
  )
  max_suppressor <- min(
    filter(
      spatial_struct_effects,
      interaction_matrix == im & effect == "suppressor"
    )$graph_median_score
  )

  im_effects <- spatial_struct_effects %>%
    filter(interaction_matrix == im) %>%
    mutate(
      max_promoter = graph_median_score == max_promoter,
      max_suppressor = graph_median_score == max_suppressor
    ) %>%
    arrange(effect)

  # Identify biggest suppressor / promoter
  table <- im_effects %>%
    kable() %>%
    kable_styling(
      latex_options = "striped"
    )

  save_kable(
    table,
    paste(
      plot_dir,
      paste0("spatial_struct_effect_table_", im, ".pdf"),
      sep = "/"
    )
  )
}
```

### Distribution of effects for each spatial structure type

Count the distribution of effects each graph type is categorized with.

```{r}
effect_counts <- spatial_struct_effects %>%
  mutate(
    effect = as.factor(effect),
    graph_type = as.factor(graph_type)
  ) %>%
  group_by(graph_type, effect) %>%
  dplyr::summarize(
    n = n()
  )

table <- effect_counts %>%
  kable() %>%
  kable_styling(
    latex_options = "striped"
  )

save_kable(
  table,
  paste(
    plot_dir,
    "spatial_struct_effect_counts.pdf",
    sep = "/"
  )
)
```

Visualize:

```{r}
# Manually set ordering for plot:
graph_ticks <- c(
  "toroidal-lattice",
  "linear-chain",
  "cycle",
  "wheel",
  "star",
  "windmill",
  "comet-kite",
  "random-barabasi-albert",
  "random-waxman"
)
graph_labels <- c(
  "Toroidal lattice",
  "Linear chain",
  "Cycle",
  "Wheel",
  "Star",
  "Windmill",
  "Comet-kite",
  "Barabasi-Albert",
  "Waxman"
)

effect_counts_fig <- effect_counts %>%
  ggplot(
    aes(
      fill = effect,
      x = graph_type,
      y = n
    )
  ) +
  geom_bar(
    position = "stack",
    stat = "identity"
  ) +
  geom_text(
    aes(label = n),
    position = position_stack(vjust = 0.5),
    size = 8,
    color = "white"
  ) +
  scale_fill_highcontrast(
    name = "Effect:",
    limits = c("suppressor", "promoter", "neither"),
    labels = c("Suppressor", "Amplifier", "Neither"),
    reverse = TRUE
  ) +
  scale_x_discrete(
    name = "Spatial Structure",
    limits = graph_ticks,
    breaks = graph_ticks,
    labels = graph_labels
  ) +
  scale_y_continuous(
    name = "Count",
    limits = c(0, 9),
    breaks = c(0, 3, 6, 9)
  ) +
  coord_flip() +
  theme(
    legend.position = "bottom"
  )

ggsave(
  paste(
    plot_dir,
    "spatial_structure_effect_distributions.pdf",
    sep = "/"
  ),
  plot = effect_counts_fig,
  width = 6,
  height = 4
)

effect_counts_fig
```

