# Graph property correlations

We screened for graph properties correlated with community transitionability scores.

## Dependencies and setup

```{r}
library(tidyverse)
library(Hmisc)
library(broom)
library(knitr)
library(kableExtra)
```

```{r}
# Check if Rmd is being compiled using bookdown
bookdown <- exists("bookdown_build")
```

```{r}
experiment_slug <- "2024-03-08-varied-interaction-matrices"
working_directory <- paste(
  "experiments",
  experiment_slug,
  "analysis",
  sep = "/"
)
# Adjust working directory if being knitted for bookdown build.
if (bookdown) {
  working_directory <- paste0(
    bookdown_wd_prefix,
    working_directory
  )
}

plot_dir <- paste(
  working_directory,
  "plots",
  sep = "/"
)

data_path <- paste(
  working_directory,
  "data",
  "world_summary_final_update_with-graph-props.csv",
  sep = "/"
)
data <- read_csv(data_path)
```

Set cowplot theme as default plotting theme.

```{r}
theme_set(theme_cowplot())
```

## Data preprocessing

```{r}
max_update <- max(data$update)
# Ensure that we just have measurements from final update.
data <- data %>%
  filter(update == max_update) %>%
  mutate(
    interaction_matrix = as.factor(interaction_matrix),
    graph_type = as.factor(graph_type),
    summary_mode = as.factor(summary_mode),
    update = as.numeric(update),
    SEED = as.factor(SEED),
    graph_file = str_split_i(DIFFUSION_SPATIAL_STRUCTURE_FILE, "/", -1)
  ) %>%
  mutate(
    graph_file = as.factor(graph_file)
  )
# write_csv(
#   data,
#   "world_summary_final_update.csv"
# )

# For each row, assign graph properties
properties <- c(
  "graph_prop_density",
  "graph_prop_degree_mean",
  "graph_prop_degree_median",
  "graph_prop_degree_variance",
  "graph_prop_girth",
  "graph_prop_degree_assortivity_coef",
  "graph_prop_num_bridges",
  "graph_prop_max_clique_size",
  "graph_prop_transitivity",
  "graph_prop_avg_clustering",
  "graph_prop_num_connected_components",
  "graph_prop_num_articulation_points",
  "graph_prop_avg_node_connectivity",
  "graph_prop_edge_connectivity",
  "graph_prop_node_connectivity",
  "graph_prop_diameter",
  "graph_prop_radius",
  "graph_prop_kemeny_constant",
  "graph_prop_global_efficiency",
  "graph_prop_wiener_index",
  "graph_prop_longest_shortest_path"
)

# (3) Pivot longer
long_data <- data %>%
  mutate(
    graph_prop_diameter = case_when(
      graph_prop_diameter == "error" ~ "-1",
      .default = graph_prop_diameter
    ),
    graph_prop_radius = case_when(
      graph_prop_radius == "error" ~ "-1",
      .default = graph_prop_radius
    ),
    graph_prop_kemeny_constant = case_when(
      graph_prop_kemeny_constant == "error" ~ "-1",
      .default = graph_prop_kemeny_constant
    )
  ) %>%
  mutate(
    graph_prop_diameter = as.numeric(graph_prop_diameter),
    graph_prop_radius = as.numeric(graph_prop_radius),
    graph_prop_kemeny_constant = as.numeric(graph_prop_kemeny_constant)
  ) %>%
  select(
    !c(
      DIFFUSION_SPATIAL_STRUCTURE_FILE,
      GROUP_REPRO_SPATIAL_STRUCTURE_FILE,
      INTERACTION_SOURCE
    )
  ) %>%
  filter(
    summary_mode == "ranked_threshold"
  ) %>%
  pivot_longer(
    cols = properties,
    names_to = "graph_property",
    values_to = "graph_property_value"
  ) %>%
  filter(
    (!is.na(graph_property_value)) & graph_property_value != "Inf" &
    (!(graph_property == "graph_prop_diameter" & (graph_property_value == "-1"))) &
    (!(graph_property == "graph_prop_radius" & (graph_property_value == "-1"))) &
    (!(graph_property == "graph_prop_kemeny_constant" & (graph_property_value == "-1")))
  ) %>%
  mutate(
    graph_property_value = as.numeric(graph_property_value),
    graph_property = str_remove(graph_property, "graph_prop_")
  ) %>%
  mutate(
    graph_property = as.factor(graph_property)
  )
# write_csv(long_data, "test.csv")
```

## Plot relationships between transitionability and graph properties

```{r}
rel_plot <- long_data %>%
  ggplot(
    aes(
      x = graph_property_value,
      y = logged_mult_score
    )
  ) +
  geom_point(aes(color = graph_type)) +
  geom_smooth(
    method = "lm",
    color = "black"
  ) +
  facet_grid(
    interaction_matrix ~ graph_property,
    scales = "free"
  )

rel_plot

ggsave(
  plot = rel_plot,
  filename = paste(
    plot_dir,
    "property_relationships.pdf",
    sep = "/"
  ),
  width = 40,
  height = 20
)

```

## Measure correlations

```{r}
# Reference for running correlations over tidy data:
# https://dominicroye.github.io/en/2019/tidy-correlation-tests-in-r/
cor_fun <- function(data) {
  cor.test(
    data$graph_property_value,
    data$logged_mult_score,
    method = "spearman",
    exact = FALSE
  ) %>% tidy()
}

nested <- long_data %>%
  select(
    c(
      interaction_matrix,
      graph_property,
      graph_property_value,
      logged_mult_score
    )
  ) %>%
  group_by(interaction_matrix, graph_property) %>%
  nest() %>%
  mutate(
    model = map(data, cor_fun)
  )

full_corr <- select(nested, -data) %>% unnest()
full_corr <- full_corr %>%
  mutate(
    abs_estimate = abs(estimate)
  ) %>%
  arrange(
    desc(abs_estimate)
  ) %>%
  group_by(
    interaction_matrix
  ) %>%
  mutate(
    p.value.adj = p.adjust(p.value, method = "holm")
  ) %>%
  filter(
    p.value.adj <= 0.05
  )

full_corr_table <- kable(full_corr) %>%
  kable_styling(latex_options = "striped")
save_kable(
  full_corr_table,
  paste(
    plot_dir,
    "correlation_table.pdf",
    sep = "/"
  )
)
```

### Top three significant correlations per interaction matrix

Break correlations down by interaction matrix

```{r}
interaction_matrices <- levels(long_data$interaction_matrix)

for (mat_type in interaction_matrices) {
  mat_data <- filter(long_data, interaction_matrix == mat_type)

  nested <- mat_data %>%
    select(
      c(
        interaction_matrix,
        graph_property,
        graph_property_value,
        logged_mult_score
      )
    ) %>%
    group_by(interaction_matrix, graph_property) %>%
    nest() %>%
    mutate(
      model = map(data, cor_fun)
    )

  im_corr <- select(nested, -data) %>% unnest()
  im_corr <- im_corr %>%
    mutate(
      abs_estimate = abs(estimate)
    ) %>%
    arrange(
      desc(abs_estimate)
    ) %>%
    ungroup() %>%
    group_by(
      interaction_matrix
    ) %>%
    mutate(
      p.value.adj = p.adjust(p.value, method = "holm")
    ) %>%
    filter(
      p.value.adj < 0.05
    )

  im_corr_table <- kable(im_corr) %>%
    kable_styling(latex_options = "striped")
  save_kable(
    im_corr_table,
    paste(
      plot_dir,
      paste0("correlation_table_", mat_type, ".pdf"),
      sep = "/"
    )
  )

  top_corr <- im_corr %>%
    slice_max(
      abs_estimate,
      n = 3
    )
  top_corr_table <- kable(top_corr) %>%
    kable_styling(latex_options = "striped")

  save_kable(
    top_corr_table,
    paste(
      plot_dir,
      paste0("t3_correlation_table_", mat_type, ".pdf"),
      sep = "/"
    )
  )

}
```

### Distrubition of correlations >= 0.5 strength

```{r}
# Look at distribution of correlations >= 0.5 strength
corr_str_thresh <- 0.5
full_corr_thresh <- full_corr %>%
  mutate(
    direction = case_when(
      estimate < 0 ~ "Negative",
      estimate >= 0 ~ "Positive"
    )
  ) %>%
  mutate(
    direction = as.factor(direction)
  ) %>%
  filter(abs_estimate >= corr_str_thresh & p.value.adj <= 0.05)

corr_counts <- full_corr_thresh %>%
  dplyr::group_by(graph_property, direction) %>%
  dplyr::summarize(
    n = n()
  )

# If something has one direction, but not other, fill in 0 for other.
# For property in properties
correlation_dirs_plot <- corr_counts %>%
  ggplot(
    aes(
      x = graph_property,
      fill = direction,
      y = n
    )
  ) +
  geom_bar(stat = "identity", position = position_dodge(), alpha = 0.75) +
  coord_flip()

ggsave(
  plot = correlation_dirs_plot,
  filename = paste(
    plot_dir,
    "most_moderate_correlations.pdf",
    sep = "/"
  )
)
correlation_dirs_plot
```